\TITLE

### 2021-01-27 final/vonLaszewski-covid.tex
start cleanup of final

%\documentclass[letterpaper, inpress]{jds} % use this for production
\documentclass[a4paper, inpress]{jds} % use this for production
%\documentclass[a4paper, review]{jds}      % use this for review

% spacing
% outcomment this for production
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\doublespace
% above needs to be commented out for production

\usepackage{subcaption}
\usepackage{calc}
\usepackage{comment}
\usepackage{todonotes}
\setuptodonotes{inline}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% the following edits should be done by Journal typesetters
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{page}{1}            % set the first page number
\jdsmonth{January}              % month
\jdsyear{2021}                  % year
\jdsvolume{xx}                  % volume number
\jdsissue{xx}                   % issue number
\jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
\jdsreceived{Jun 30, 2020}       % optional: comment it out if no received date
\jdsaccepted{???, 2021}         % optional: comment it out if no accepted date
%% manually set running header for a shorter list of authors if needed
% \shortauthors{A Author A, et al.}

@@

%\linelabel{comment:level}

@@

\begin{comment}
\begin{question} \normalfont What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
\label{q:4}
\end{question}
\end{comment}

@@

\begin{comment}
Further, the obtained sorted order of the best prediction models by
minimum error based on the risk factors for cases and deaths are as
follows:

\begin{description}

\item[Order for cases sorted by minimum model error:] 
  POP\_DENSITY\_2010, PHLTH, BLACK, HISPANIC, NHOSP, INSURANCE, 
  CHOLSCREEN, DIABETES, STROKE, CHD, CHECKUP, NBEDS, SVI\_OVERALL, 
  KIDNEY, CASTHMA, BPMED, LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, 
  MHLTH, COPD, SENIOR, CANCER, NBEDS/1000, BPHIGH, NONE, HIGHCHOL, 
  BINGE, OBESITY, SVI\_MINORITY

\item[Order for deaths sorted by minimum model error:] 
  SVI\_OVERALL, CSMOKING, HISPANIC, NONE, POP\_DENSITY\_2010, 
  SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, MHLTH, LPA, 
  POVERTY, OBESITY, CANCER, NBEDS/1000, BLACK, DIABETES, ARTHRITIS, 
  NBEDS, NHOSP, BPHIGH, PHLTH, CHECKUP, CHD, STROKE, HIGHCHOL, 
  ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY

\end{description}
\end{comment}

@@

\begin{figure}[!htbp]
%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Cases.pdf}
%        \vspace{-1cm}
%        \caption{Best predictions for cases over all risk factors.}
%        \label{fig:place-top1-cases}
%
%    \end{minipage}
%    \ \
%    \begin{minipage}{0.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Death.pdf}
%        \vspace{-1cm}
%        \caption{Best predictions for deaths over all risk factors.}
%        \label{fig:place-top1-death}
%    \end{minipage}

%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Cases.pdf}
%        \vspace{-1cm}
%        \caption{Top 5 predictions from all cases over all risk factors.}
%        \label{fig:place-top5-cases}
%    \end{minipage}
%    \ \
%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Death.pdf}
%        \vspace{-1cm}
%        \caption{Top 5 predictions from all deaths  over all risk factors.}
%        \label{fig:place-top5-death}
%    \end{minipage}
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW3.pdf}
        \caption{Cumulative error for deaths.}
        \label{fig:place-top10-death}
    \end{subfigure}
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW3.pdf}
        \caption{Cumulative error for cases.}
        \label{fig:place-top10-cases}
    \end{subfigure}

@@

\begin{comment}

\subsection{Impact and Design of a Multi-Covariate-LSTM}
\label{sec:lstm-many}

In our next comprehensive analysis, we try to answer Question
\ref{q:4} if all risk factors impact the analysis. Thus we feed the
covariates into the recurrent network as additional features. In
addition to cases and deaths, this leads to 35 variables, including 33
unique covariates. We looked at two related ways of doing this: (a)
have covariates fixed in time (static), and (b) multiply the
covariates by the times series (we used the average of cases and
deaths)

We expected the second choice to be the most natural as it gives all
35 input streams similar time dependence. However, in our initial
tests, we found these choices gave very similar results, and so we
focused on the simpler choice (a) in this paper. In the experiments
discussed here, we either used 0, 1, or 33 covariates. Choosing 33
gives the best description (reducing loss by about 1\%) while using
each covariate as a solitary extra feature allows us a rather
sensitive study of the importance of each individual covariate as we
noted in Section \ref{sec:lstm-covariate}.

The final choice was the quantities to predict and train on. We choose
this to be both the cases and deaths values at the time value
following the sequence combined with the same quantities for the
following 14 days. This gives us 30 different output features for each
sequence. We renormalized, so the two basic features (following day)
contributed 50\% of the loss function. We supported this in TensorFlow
with a custom mean square error (MSE) loss function that dropped
training values that were not available for time sequences near the
end of the examined period. While one can not train on future values,
we can predict them, as seen in our sample outputs. We extensively
studied both 2 and 30 output features, and they give similar
descriptions of the observed data with the 30 feature choice enabling
14-day predictions.

We performed an extensive hyper-parameter search and we show the
values in Table \ref{tab:model} that give robust results without
overfitting. Furthermore, we looked at four different activation
functions for this analysis in the LSTM and fully connected
layers. Scaled Exponential Linear Unit (SELU), REctified Linear Unit
(RELU) and TANH activations gave very similar results while SIGMOID
activation was much worse. However we still use a SIGMOID function in
the last stage just as we did in Section \ref{sec:lstm-covariate}. We
also looked at the number of hidden units (separately for LSTM and
fully connected layers), the number of LSTM layers and the dropout
values.  Only the latter produced a significant change and, for
example, removing dropouts decreased the loss by about 16\% but we
retained them to provide a robust environment. We chose randomly 20\%
of the data for validation and testing and this had similar loss
values to the training data (the other 80\%) with our chosen dropout.

The final technical issue concerns the data normalization where each
feature is transformed by a feature dependent -- but city independent
-- way to lie between 0 and 1. Further all extensive (proportional to
size) covariates were first divided by the city population except for
the population itself where we took the logarithm; intensive
covariates were just rescaled to lie between 0 and 1. The basic
COVID-19 daily data is extensive but compared two approaches with and
without renormalization, but in this case, we took the square root of
the data value.  This is natural if there are counting (square root(N)
for N observations) errors as the deep learning loss function is
$(predicted-observed)^{2}$ without the error traditional in
$\chi^{2}$, namely $((predicted-observed)/error)^{2}$. This approach
with standard deep learning MSE appears sound instead of dividing by
the population. That will emphasize the error contribution of smaller
cities but is still interesting to look at due to the presence of many
systematic errors.

\section{Direct Deductions from Deep Learning Data Representation}

The deep learning model contains a lot of information much of which we
have already analyzed and discussed in Section
\ref{sec:lstm-covariate} and Figures \ref{fig:magic-1} and
\ref{fig:error-case-popdensity}. In addition, we have looked at the
sensitivity of the results as measured by the predicted daily rates
and the value of the loss function. In Table \ref{tab:corr-matrix}, we
present one way of looking at it. We calculate the correlation
coefficients of many of the covariates from an accurate LSTM model fit
to the data. We divide the time series into three regions

\begin{itemize}
\item \textbf{Past} or all days up to 14 days before the final
  observed date (May 25, 2020)
\item \textbf{Now} or the 14 days from May 12 to May 25, 2020
\item \textbf{Future} or the 14 days after May 25, 2020 predicted by
  the model
\end{itemize}

These results are different between population normalized or direct
non-normalized data. We have kept the square-root used in the model
but divided the model values by the square root of the population when
computing the correlation. Significant positive correlations are seen
for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP ,
while negative correlations for PVI and NORM\_POP. We also looked at
the impact of these covariates on the evolution function. As this
operates on data at previous time values, this is sensitive to
variations across cities and times. The evolution operator is
intuitively the time derivative of the absolute values of daily rates
seen in the correlation in Table 3. Nhosp and Estbeds have the most
significance in impacting (as a reduction) the values of both the
infection and fatality rates from the evolution operator. The fatality
rates are also significantly impacted by INSURANCE and NORM\_POP
(reduction) and SENIOR and OBESITY (increase) from the evolution
operator. NORM\_POP impacts infection values as an increase.  We can
also identify which covariates have the most significant impact on the
loss function. Here NHOSP, CHD, KIDNEY, BINGE, and PHLTH give 5\% or
more reduction in loss function while NORM\_POP, SVI\_OVERALL,
CASTHMA, DIABETES, and BPHIGH have small (1\%) effects.

\begin{table}[!h]
\caption{Correlation Matrix}
\bigskip
\label{tab:corr-matrix}
\begin{center}
\begin{tabular}{lrrrrrr}
\toprule
Risk Factor		&
\makecell{Past\\Case} &
\makecell{Now\\Case} &
\makecell{Future\\Case} &
\makecell{Past\\Death} &
\makecell{Now\\Death} &
\makecell{Future\\Death} \\
\midrule
PVI                     & -0.425      & -0.292      & -0.124      & -0.434      & -0.373      & -0.239    \\
INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
NBEDS                   & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
NBEDS/1000              & 0.138       & 0.121       & 0.209       & 0.092       & 0.069       & 0.088     \\
NHOSP                   & 0.074       & 0.119       & 0.165       & -0.009      & 0.028       & 0.021     \\
ESTBEDS                 & 0.136       & 0.131       & 0.217       & 0.089       & 0.080        & 0.109     \\
SENIOR                  & 0.104       & 0.028       & 0.033       & 0.152       & 0.164       & 0.156     \\
BLACK                   & 0.209       & 0.099       & 0.208       & 0.216       & 0.085       & 0.116     \\
HISPANIC                & -0.072      & -0.032      & -0.064      & -0.091      & -0.068      & -0.046    \\
POP\_DENSITY\_2010      & -0.302      & -0.210      & -0.175      & -0.272      & -0.231      & -0.216    \\
POVERTY                 & 0.128       & 0.059       & 0.200       & 0.148       & 0.031       & 0.080     \\
SVI\_MINORITY           & 0.029       & 0.069       & 0.081       & 0.010        & 0.006       & 0.087     \\
SVI\_OVERALL            & 0.140       & 0.158       & 0.238       & 0.125       & 0.113       & 0.179     \\
CASTHMA                 & 0.238       & 0.271       & 0.307       & 0.205       & 0.288       & 0.309     \\
HIGHCHOL                & 0.202       & 0.185       & 0.131       & 0.192       & 0.243       & 0.224     \\
DIABETES                & 0.200       & 0.200     & 0.178         & 0.206       & 0.232       & 0.230     \\
OBESITY                 & 0.199       & 0.205       & 0.194       & 0.191       & 0.229       & 0.212     \\
CANCER                  & -0.145      & -0.102      & -0.129      & -0.138      & -0.058      & -0.059    \\
STROKE                  & 0.123       & 0.125       & 0.117       & 0.167       & 0.143       & 0.135     \\
MHLTH                   & 0.110       & 0.161       & 0.186       & 0.076       & 0.175       & 0.222     \\
CSMOKING                & 0.189       & 0.258       & 0.249       & 0.160        & 0.289       & 0.323     \\
CHOLSCREEN              & 0.164       & 0.001       & -0.026      & 0.187       & 0.070       & 0.038     \\
INSURANCE               & 0.018       & 0.017       & 0.012       & 0.018       & -0.002      & 0.001     \\
CHD                     & 0.225       & 0.284       & 0.233       & 0.178       & 0.327       & 0.337     \\
CHECKUP                 & 0.206       & 0.221       & 0.280       & 0.201       & 0.272       & 0.343     \\
KIDNEY                  & 0.045       & 0.028       & 0.032       & 0.022       & 0.071       & 0.053     \\
BINGE                   & 0.031       & 0.044       & 0.009       & 0.083       & 0.030       & 0.022     \\
LPA                     & 0.191       & 0.187       & 0.197       & 0.147       & 0.220       & 0.214     \\
ARTHRITIS               & 0.046       & 0.013       & -0.021      & 0.061       & 0.089       & 0.079     \\
BPMED                   & 0.060       & 0.080        & 0.124       & 0.056       & 0.073      & 0.090      \\
PHLTH                   & 0.064       & 0.132       & 0.104       & 0.022       & 0.129       & 0.114     \\
BPHIGH                  & 0.056       & 0.051       & 0.048       & 0.058       & 0.084       & 0.068     \\
COPD                    & 0.163       & 0.163       & 0.153       & 0.175       & 0.220       & 0.240     \\
NORM\_POP               & -0.305      & -0.258      & -0.218      & -0.193      & -0.218      & -0.099  \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\end{comment}

@@

\begin{comment}
Furthermore, we are currently conducting experiments with multiple
risk factors at the same time to identify the most significant
combinations of them.  As we have shown that the bivariate analysis
has a locality dependency, we will explore categorizing locations with
similar demographics and conduct further experiments on demographical
similar regions.
\end{comment}

>>>>

\documentclass[a4paper, inpress]{jds} % use this for production

\usepackage{subcaption}
\usepackage{calc}
\usepackage{todonotes}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% the following edits should be done by Journal typesetters
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{page}{1}            % set the first page number
\jdsmonth{January}              % month
\jdsyear{2021}                  % year
\jdsvolume{xx}                  % volume number
\jdsissue{xx}                   % issue number
\jdsdoi{xx.xxxx/xxxxxxxxx}      % doi
\jdsreceived{June 30, 2020}       % optional: comment it out if no received date
\jdsaccepted{January 23, 2021}         % optional: comment it out if no accepted date
%% manually set running header for a shorter list of authors if needed
% \shortauthors{A Author A, et al.}

@@

@@

@@

@@

\begin{figure}[!htbp]
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_DeathNEW3.pdf}
        \caption{Cumulative error for deaths.}
        \label{fig:place-top10-death}
    \end{subfigure}
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_CasesNEW3.pdf}
        \caption{Cumulative error for cases.}
        \label{fig:place-top10-cases}
    \end{subfigure}

@@

@@
